On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-19 paper/paper3d.tex
reg update

\begin{abstract}
 Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by: 
\begin{inparaenum}[1)] 
\item 
cooling, which reduces pressure support; and/or
\item 
viscosity, which reduces rotational support. 
\end{inparaenum}
We analyse the gravitational stability of accretion disks
including these non-ideal effects explicitly.   
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. We suggest that realistic disk fragmentation can be
physically attributed,  
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%alternative abstract

@@

\subsection{Viscosity removes rotational support}
%there is also viscous damping which stabilizes
Viscous disks can develop GI even when $Q>1$. This is because
viscosity compromises rotational support against self-gravity for
long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect  
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism for planetesimational formation %frag of dust
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

@@

Experimentally, however, it is found that if $\beta$ ($\alpha$) is
too small (large), 
\begin{align}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. 
There is considerable debate on the exact value of $\beta_\mathrm{c}$
and whether or not a critical cooling time can 
be defined at all \citep{meru11,meru12,paardekooper12}. However, it is
generally accepted that \emph{steady, gravito-turbulent disks do not
  exist for sufficiently rapid cooling}.

>>>>

\begin{abstract}
 Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by: 
\begin{inparaenum}[1)] 
\item 
cooling, which reduces pressure support; and/or
\item 
viscosity, which reduces rotational support. 
\end{inparaenum}
We analyse the gravitational stability of accretion disks
including these non-ideal effects explicitly.   
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. We suggest that realistic disk fragmentation can be
physically attributed,  
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%we suggest turbulent stresses play an active, dynamical role in disk fragmentation by 
%removing rotational stabilization against self-gravity. 
%in the sense of small-scale transport

@@

\subsection{Viscosity removes rotational support}
%there is also viscous damping which stabilizes
Viscous disks can develop GI even when $Q>1$. This is because
viscosity compromises rotational support against self-gravity for
long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect  
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism to enhance particle clumping for planetesimational formation 
%frag of dust 
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

@@

Experimentally, however, it is found that if $\beta$ ($\alpha$) is
too small (large), 
\begin{align}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. 
There is considerable debate on the exact value of $\beta_\mathrm{c}$
and whether or not a critical cooling time can 
be defined at all \citep{meru11,meru12,paardekooper12,hopkins13}. However, it is
generally accepted that \emph{steady, gravito-turbulent disks do not
  exist for sufficiently rapid cooling}.

### 2016-01-19 paper/summary.tex
reg update

The effect of cooling and irradiation on GI is quantified by the
dispersion relation 
Eq. \ref{inviscid}. The sufficient condition for stability 
depends on the irradiation level (Eq. \ref{stable_condition}), and 
\emph{independent} of the cooling rate, provided it is non-zero and
finite. This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\lesssim O(1)$.

@@

We generalize the `viscous gravitational instability', previously 
studied in isothermal disks
\citep{lynden-bell74,willerding92,gammie96}, to include 
cooling, viscous heating and irradiation. Our fiducial example,  
where the viscosity and self-gravity are inversely
related ($\alpha\propto Q^{-2}$), is a toy model for a
gravito-turbulent disk. We find viscous GI occurs on orbital  
timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
of a maximum stress sustainable by gravito-turbulence established in
previous numerical simulations. Because viscosity affects
rotational support, viscous GI occurs on large scales, $kH\gtrsim
O(1)$.  Furthermore, irradiation preferentially
stabilizes small-scale perturbations.

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence
\citep{rafikov15}. We show that, for linear dynamics, realistic 
cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
actually modeled by `beta' cooling with a  
non-zero irradiation parameter ($\theta\neq0$). However, 
numerical simulations typically employ beta cooling with 
$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
are such simulations to realistic PPDs must then be further 
investigated. The important implication of having 
$\theta\sim 1$ in a PPD is that cooling alone cannot 
destabilize the disk at large radii. Instead, viscous GI occurs on
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. By contrast, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

We note that power in gravito-turbulence is dominated by scales with
$kH\sim 1$ \citep{cossins09}, and this is always  
well-resolved by modern numerical simulations. However, this does not
necessarily imply small-scale dynamics are unimportant, as is evident
from the non-converging simulations described above.

@@

\subsection{Outstanding issues}
%the link between the linear GI's discussed in this paper 
%and frag 
True disk fragmentation is a non-linear process characterized by clumps   
reaching densities orders of magnitude above the ambient value, and
which survives tidal shear and shocks \citep{paardekooper12}.  
%cite the young papers 
Clearly, our linear models cannot address fragmentation
directly. Technically, we have only demonstrated that a 
gravito-turbulent state becomes dynamically unstable, and thus should
not persist, when cooling (viscosity) is too rapid (large).  
However, numerical simulations show that there are only two
possible outcomes of cooling, self-gravitating disks: steady
gravito-turbulence or fragmentation. It seems then reasonable to 
speculate that the non-existence of a gravito-turbulent state, here due to
dynamical instability, implies disk fragmentation.

@@

>>>>

The effect of cooling and irradiation on GI is quantified by the
dispersion relation 
Eq. \ref{inviscid}. The sufficient condition for stability 
depends on the irradiation level (Eq. \ref{stable_condition}), and 
\emph{independent} of the cooling rate, provided it is non-zero and
finite. This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We generalize the `viscous gravitational instability', previously 
studied in isothermal disks
\citep{lynden-bell74,willerding92,gammie96}, to include 
cooling, viscous heating and irradiation. In our fiducial disk,  
where viscosity and self-gravity are inversely
related ($\alpha\propto Q^{-2}$) to model a
gravito-turbulent background, viscous GI occurs on orbital  
timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
of a maximum stress sustainable by gravito-turbulence established in
previous numerical simulations \citep{rice05}. Because viscosity affects 
rotational support, viscous GI occurs on large scales, $kH\lesssim
O(1)$.  Furthermore, irradiation preferentially
stabilizes small-scale perturbations. %maybe that's why irradiated disks produce smaller alpha

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence
\citep{rafikov15}. We show that, for linear dynamics, realistic 
cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
actually modeled by `beta' cooling with a  
non-zero irradiation parameter ($\theta\neq0$), in contrast to 
numerical simulations which typically employ beta cooling with  
$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
are such simulations to realistic PPDs must then be further 
investigated.

The important implication of having 
$\theta\sim 1$ in a PPD is that cooling alone cannot 
destabilize the disk at large radii. Instead, viscous GI occurs on
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

We note that power in gravito-turbulence is dominated by scales with
$kH\sim 1$ \citep{cossins09}, and this is always  
well-resolved by modern numerical simulations. However, this does not
necessarily imply small-scale dynamics are unimportant \citep[especially for non-linear evolution, see][]{young15}, 
as is evident from the non-converging simulations described above.

@@

\subsection{Outstanding issues}
%the link between the linear GI's discussed in this paper 
%and frag 
True disk fragmentation is a non-linear process characterized by clumps   
reaching densities orders of magnitude above the ambient value, and
which survives tidal shear and shocks \citep{paardekooper12,young16}.  
%cite the young papers 
Clearly, our linear models cannot address fragmentation
directly. Technically, we have only demonstrated that a 
gravito-turbulent state becomes dynamically unstable, and thus should
not persist, when cooling (viscosity) is too rapid (large).  
However, numerical simulations show that there are only two
possible outcomes of cooling, self-gravitating disks: steady
gravito-turbulence or fragmentation. It seems then reasonable to 
speculate that the non-existence of a gravito-turbulent state, here due to
dynamical instability, implies disk fragmentation.

{\bf comment on stochasitc frag. we only consider 
how to enhance GI by reducing rotation and pressure. but a third route is 
increasing sigma. according to toomre Q these are the 3 effects one should consider
for SG. turbulent flucations may occasionally produce high sigma dens perts. 
possible relation to current framework? high sigma perts arise because of viscous GI? 
high alpha $\to$ small-scale transport $\to$ enhance over-dense regions}

@@

%large stresses -> viscous GI operates quickly -> get non-linear den perts quickly
